Moral Foundations of U.S. Political News Organizations

### 2019-04-15 pilot_manuscript.Rmd
DONE!!!!!!

abstract: |
  The media ecosystem has grown, and political opinions have diverged such that there are competing conceptions of objective truth. Commentators often point to political biases in news coverage as a catalyst for this political divide. The Moral Foundations Dictionary (MFD) facilitates identification of ideological leanings in text through frequency of the occurrence of certain words. Through web scraping, the researchers extracted articles from popular news sources' websites, calculated MFD word frequencies, and identified words' respective valences. This process attempts to uncover news outlets' positive or negative endorsements of certain moral dimensions concomitant with a particular ideology. In Experiment 1, the researchers gathered poltical articles from four sources. They were unable to reveal significant differences in moral or political endorsements, but they solidified the method to be employed in further research. In Experiment 2, the researchers will expand their number of sources to 10 and will analyze articles that pertain to the 2018 confirmation hearings of U.S. Supreme Court Justice Brett Kavanaugh. This topic was selected due to the moral disputes associated with his nomination.

@@

@Suhler2011 also criticized the content and taxonomy of the five foundations. The authors criticize @Haidt2007's omissions of additional foundations, including *industry* and *modesty*, claiming these concepts are moralized in many societies worldwide. Likewise, the authors question whether or not the foundations are sufficiently distinct as to stand as their own foundation. For example, @Suhler2011 posit that *ingroup/loyalty* is merely a group-focused version of *harm/care*. Work by @Graham2011 might serve to rebut this criticism, as the researchers found their original five-factor structure seemed to best fit the data when validating the Moral Foundations Questionnaire. Finally, @Suhler2011 point out that particular concepts related to a foundation, including "anger" in *fairness/reciprocity* and "deception" in *ingroup/loyalty*, could be ascribed to any of the other foundations as well. In other words, it becomes difficult to recognize a particular concept as indidcative of any one foundation when, in theory, they could be applied to all five [@Haidt2007; @Suhler2011].

These criticisms of Moral Foundations Theory are valid and should be taken into consideration when conducting research with instruments derived from MFT. The current authors argue these criticisms are especially valid when considered alongside questions regarding the Moral Foundations Dictionary they state herein. However, there exists compelling evidence regarding the validity of Moral Foundations Theory, albeit regarding its application solely through the Moral Foundations Questionnaire. This evidence is discussed within a brief explanation of the Moral Foundations Questionnaire and its relationship to the Moral Fondations Dictionary.

@@

The Moral Foundations Questionnaire has been validated by multiple researchers. Likewise, its five-factor structure has been demonstrated to fit data in multiple countries [@Graham2011; @Davies2014]. In their article introducing the 30-item MFQ, @Graham2011 conducted an exploratory factor analysis (EFA) and found that a five-factor model fit better than a one, two, or three-factor model. @Davies2014 conducted a confirmatory factor analysis (CFA) of the MFQ with a sample from New Zealand and likewise found the five-factor model provided the best fit. While @Davies2014 concede the US and New Zealand share many similarities as Western nations, which could raise questions regarding the validity of the MFQ in non-Western nations. However, there are striking differences between the two countries, including the lack of a two-party political system in New Zealand, that provide grounds for claiming the MFQ generalizes beyond the United States. Furthermore, @Graham2011 claim to find a "reasonable" degree of generalizability for the MFQ across participants from many different regions in the world. These two bodies of work also provide the best available evidence that the five moral foundations are sufficiently distinct from one another, though broader criticisms of MFT raised by @Suhler2011 should still be taken into account in studies involving the theory.

@@

While these concerns regarding the replication of social priming studies are valid and deserve further investigation, @Druckman2004 does not purport to demonstrate a widespread effect of social priming on the American electorate. In other words, this reseach makes no claim to empirically supported priming effects. Rather, @Druckman2004 chronicle the efforts on the part of the Nixon Administration to prop up the president's supposed strengths while downplaying his weaknesses. These tactics were deployed through the careful use of language in order to achieve the administration's political goals. As such, @Druckman2004's research on Nixon serves as an example of language's potential utility in the propogation of desirable political opinions. The researcher's investigation of news media's focus on specific issues during the Carter Administration likewise provide an example of language as a potential conduit for the transfer of politically biased information. The idea that even 1970s news media could contain political biases is of particular interest to the current study, which investigates similar phenomena in contemporary news media.

@@

For Experiment 1, the researchers approached the study with the intention to answer a method question. That is, this portion of the current research was conducted in order to solidify the best method by which to analyze political news text under the Moral Foundations Theory framework while also alleviating some of the aforementioned valence problem observed in the Moral Foundations Dictionary. The researchers hypothesized the news sources genrally perceived as liberal leaning (*NPR* and *The New York Times*) would contain MFD words and valences indicating endorsements of the individualizing moral foundations (*harm/care* and *fairness/reciprocity*). Additionally, the researchers hypothesized the two sources generally perceived to be conservative leaning (*Fox News* and *Breitbart*) would feature MFD words and valences indicating equal endorsement of all five foundations.

@@

Political articles were collected from the websites of four notable U.S. news sources, a process known as web scraping. The sources were *The New York Times*, *National Public Radio (NPR)*, *Fox News*, and *Breitbart*. They were selected for their widespread recognition and the fact politcial partisans have strong preferences for some sources over others. The researchers determined the political lean of each source by referencing @Mitchell2014's article demonstrating the self-reported idealogical consistency represented by the consumers of several news sources. In general, *The New York Times* and *NPR* are preferred by consumers reporting a liberal bias or lean. In contrast, *Fox News* and *Breitbart* are believed to have a conservative bias or lean. @Mitchell2014's article presented political ideology as a scale ranging from "consistently liberal" to "consistently conservative." In between these extremes lie more moderate positions, including "mostly liberal," "mixed," and "mostly conservative." Owing to the lower number of sources analyzed herein, the researchers elected to categorize the sources as either "liberal" and "conservative" in order to form a basis for comparison.

@@

The source article words were compiled into a dataset where they were matched up with their counterparts in the MFD along with their valence and a percentage of their occurrence. Therefore, for each article, the percentage of the number of *harm/care* words occurring in the articles were calculated, and this process was repeated for each of the foundations. Words' percent occurence were multiplied by their *z*-scored valence. Valences were *z*-scored in order to eliminate any ambiguity regarding the direction of the valence. Positive values indicate positive valence, and negative values indicate negative valence. Words were categorized in accordance to their MFD affiliation, creating a weighted sum for each moral foundation.

@@

As seen in Table \@ref(tab:exp1-source-descriptives), the sources are similar in some aspects yet different in others. Valence appears to be slightly positive across all sources. The standard deviations seem to indicate little to no presence of a difference in valence. 
*The New York Times* published the greatest number of articles as well as total words. *Breitbart* featured the lowest number of articles, and *NPR* the lowest number of total words from all articles. Per indvidual article, however, *Breitbart* appears to feature the highest average number of words as well as unique words. Once again the standard deviations call into question any apparent differences between sources. Finally, *Fox News* appears to be the most readable source on average, while *The New York Times* is the least readable. This might be attributable to the greater number of tokens in the average *New York Times* article compared to *Fox News*. As before, the standard deviations for readability hamper one's ability to draw solid conclusions regarding differences in the sources' use of language.

@@

Owing to the exploratory nature of Experiment 1, the researchers were afforded the opportunity to consider changes to the method to be utilized in Experiment 2. Generally speaking, the researchers believe their methodology to be sound. Web scraping methods and text processing remain viable methods for collecting large amounts of text and subsequently rendering that text in a form suitable for data analysis. Experiment 1 also demonstrated a method by which to address inherent problems in the Moral Foundations Dictionary relating to valence. The solution provided in Experiment 1 appears to provide insights into the MFD words where none previously existed. Finally, calculating weighted percentages and sums for each moral foundation provides an easily interpreted summary of MFD word positivity and occurence.

@@

The government shutdown ended 35 days later when Congress passed a continuing resolution to fund the government for three weeks while further negotiations regarding the budget would take place. President Trump signed the bill, ending the longest govenrment shutdown in U.S. history [@Axelrod2019]. Many political commentators cite the mounting pressure on Trump and Congress stemming from significant costs both to unpaid American workers and the economy as a whole [@Kheel2019]. Perhaps owing to the financial pressures experienced by thousands of federal employees and their families as a result of a political quarrel, President Trump saw his approval rating fall as the shutdown progressed [@Peoples2019]. While the shutdown and its associated burdens were unpopular, the issues discussed concurrent to the shutdown offer the potential for more divergent opinions.

@@

In contrast to Experiment 1, the researchers approached Experiment 2 with the intention to confirm the method employed was valid for the analysis of the scraped text as well as for any inferences drawn from the analyses. For Experiment 2, the researchers hypothesized that news sources perceived as liberal will exhibit positive endorsements of the individualizing moral foundations (*harm/care* and *fairness/reciprocity*) in their articles reporting on both the Kavanaugh confirmation hearing as well as the 2018-2019 government shutdown. News sources perceived as conservative are hypothesized to positively endorse all five foundations equally in their coverage of the Kavanaugh hearing and the government shutdown. The researchers tested the hypothesis by analyzing the content scraped from news sources' web pages spanning the two weeks before (September 13, 2018) and two weeks after (October 11, 2018) Kavanaugh's confirmation hearing, owing to its prominence in the news. Likewise, the researchers analyzed content spanning two weeks before the start of the government shutdown (December 8, 2018) to two weeks following the end of the shutdown (February 8, 2019). The content will be analyzed for valence and moral alignment under Moral Foundations Theory.

@@

The same process was followed for scraping articles related to the partical U.S. Government shutdown of 2018-2019. The articles scraped were published between December 8, 2018 and February 8, 2019 inclusive. Once again, the researchers elected to scrape articles published two weeks before and after the event in question in order to capitalize on the shutdown's saturation in American news media.

@@

As in Experiment 1, the text was scanned with *ngram*. Again, blank articles, text from the Disqus system, and duplicate articles were removed [@Schmidt2017]. The text was processed and stemmed in order to convert to a usable form for further analysis [@Feinerer2017]. Words were subsequently matched with their valences from @Warriner2013. Depending on the results of @Jordan2019's research, alternative forms of the Moral Foudnations Dictionary may be imported instead of the original dictionary.

Using the *tm* and *ngram* packages in *R*, the researchers processed the text in order to convert it to lowercase, fix spacing anomalies, and remove punctuation [@Feinerer2017]. Each individual word was reduced to its stem (i.e., *diseased* was stemmed to *diseas*). Once again, the same procedure was applied to the MFD words and the words in the @Warriner2013 dataset. Using the @Warriner2013 dictionary, the words in the MFD were assigned their respective valence. The researchers obtained the words' percent occurence in the text. Once again, percents were multiplied by *z*-scored valence and categorized into their proper MFD category.

@@

Table \@ref(tab:exp2-source-descriptives-kav) displays the descriptive statistics for sources' writing on the Kavanaugh confirmation hearing. The sources were similar in most basic linguistic aspects, except for number of articles. Sources varied on the number of articles published pertaining to this topic. For example, *The Huffington Post* appears to have published only 27 aritcles while *NPR* published 757 articles. Valence was found to be slightly positive across all sources with the standard deviations precluding any conclusions supporting the presence of differences. *Slate* produced the most total words with the most tokens on average. *Rush Limbaugh* featured the highest number of types on average. *The Blaze* was the most readable source on average by grade level while *Fox News* was the least readable. The standard deviations for these statistics, however, preclude conclusions regarding differences in the sources' use of language.

Table \@ref(tab:exp2-source-descriptives-gs) displays descriptive statistics for articles about the partial government shutdown. Like the Kavanaugh hearing, the sources were similar in average valence (slightly positive). Once again, there was variation in the number of articles published by each sorce on this topic. *The Hugginton Post* and *The Blaze* published fewer than 100 articles while *Slate* published 1,013 articles. *Slate* again featured the most total words and mean tokens. *Rush Limbaugh* again had the most types on average. For this topic, *Slate* was the most readable while *The Huffington Post* was the least readable. For each statistic, the standard deviations render any assertions regarding linguistic differences inconclusive, at least on a descriptive level.

@@

To analyze if news sources adhered to differences in word use based on their target audience, we utilized a multilevel model (MLM) to analyze if news sources leveraged different vocabularies based on target audience. Once again, the researchers used the *nlme* library in *R* [@Pinheiro2017]. Each foundation's weighted percentage was predicted by the source's political lean, using the individual source as a random intercept to control for the nested ata structure. MLMs were constructed from datasets compiled for both topics of interest: the Kavanaugh hearing and the partial government shutdown of 2018-2019.

For the Kavanuagh topic, the multilevel model indicated the presence of a significant effect for *harm/care*, but the practical effect denoted by Cohen's *d* was found to be small. There were no other significant or practical effect of political lean for any of the other four moral foundations. The effect for *harm/care* was in the hypothesized direction with liberal sources tending to use more positively *harm/care* words than consevative sources. Descriptive and test statistics, *p*-values and effect sizes (Cohen's *d*) can be found in Table \@ref(tab:exp2-tablekav).

For news articles about the partial U.S. Federal Government Shutdown of 2018-2019, there were no significant or practical effects of political lean for the moral foundations. A small-to-medium effect size was observed for *authority/respect*. The effect was in the predicted direction as well, as conservative sources tended to offer more positive endorsements of the foundation than liberal sources. This is similar to Experiment 1 in which the largest effect size was observed for *authority/respect*. As noted before, the effect found in Experiment 1 was in the opposite direction as what was hypothesized. Thus, it is difficult to draw comparisons between the two studies despite the similar pattern for effect size. Owing to a lack of matching effects for either Experiment 1 or the Kavanaugh topic, there is doubt as to whether or not a practical or generalized effect exists for *authority/respect*.

@@

The results obtained in Experiment 2 did not completely confirm either of the hypotheses. The researchers found little compelling evidence of an effect of partisan lean on MFD endorsement. The only significant effect was found for *harm/care* in the Kavanuagh confirmation hearing dataset. While the effect was in the hypothesized direction (higher positive endorsement for liberal sources), the effect size was small. While there were no significant effects found for the government shutdown topic, a small-to-medium effect size was calculated for the *authority/respect* foundation. Once again, the effect was in the hypothesized direction with conservative sources providing higher postive endorsements of the foundation relative to liberal sources.

Both of these effects are consistent with both the resesarch hypotheses as well as previous findings in Moral Foundations Theory scholarship. It should be noted these effects were found in isolation, and similar patterns were not found across the other studies. These findings undermine any inferences of political bias stemming from the content of news articles. The results obtained for Experiment 2 call into question the Moral Foundations Dictionary's efficacy as a tool for establishing differences between political news sources by partisan lean.

@@

Within the theoretical framework of Moral Foundations Theory [@Haidt2007], the researchers attempted to devise a method leveraging the Moral Foundations Dictionary [@Graham2009] in order to quantify political bias stemming from content published by several prominent American news sources. This method was carried out in several parts. First, the researchers collected the articles through web scraping techniques in *R* [@Wickham2016]. Second, the text was processed using *ngram* and *tm* in *R* to render it in a form conducive to computerized statistical analysis [@Feinerer2017]. Third, percent occurrence of words in the Moral Foundations Dictionary (MFD) was calculated on each source in order to establish how often words indicating endorsement of the five moral foundations appeared. Frequency of MFD word occurrence per article was weighted for valence by multiplying the scores by mean *z-scored* valence ratings from @Warriner2013. Finally, the researchers constructed multi-level models using *nlme* in *R* [@Pinheiro2017] for each foundation in which weighted percentages were predicted by each source's political lean. Individual source was held as a random intercept to control for correlated error. This process was undertaken once in Experiment 1 for general political news and twice in Experiment 2 using articles pertianing to two specific topics in American news: Brett Kavanaugh's confirmation to the Supreme Court and the most recent (as of this writing) partial U.S. Government shutdown from December 2018 to January 2019.

@@

For articles pertaining to the Kavanaugh hearing, there was a statistically significant effect for *harm/care*, but the effect size was modest. No other statistically or practically significant effects were found for the Kavanaugh topic. Likewise, there were no statistically or pracitically signifcant effects revealed for the government shutdown topic. However, there was a small-to-medium effect size for *authority/respect*. The fact these patterns do not appear in any of the other studies undertaken herein decreases the researchers' ability to draw any valuable conclusions regarding the MFD's ability to reveal political bias in American news media.

>>>>

abstract: |
  The media ecosystem has grown, and political opinions have diverged such that there are competing conceptions of objective truth. Commentators often point to political biases in news coverage as a catalyst for this political divide. The Moral Foundations Dictionary (MFD) facilitates identification of ideological leanings in text through frequency of the occurrence of certain words. Through web scraping, the researchers extracted articles from popular news sources' websites, calculated MFD word frequencies, and identified words' respective valences. This process attempts to uncover news outlets' positive or negative endorsements of certain moral dimensions concomitant with a particular ideology. In Experiment 1, the researchers gathered poltical articles from four sources. They were unable to reveal significant differences in moral or political endorsements, but they solidified the method to be employed in further research. In Experiment 2, the researchers will expanded their number of sources to 10 and analyzed articles that pertain to two specific topics: the 2018 confirmation hearings of U.S. Supreme Court Justice Brett Kavanaugh and the partial U.S. Goverment Shutdown of 2018-2019. Once again, no significant differences in moral or political endorsements were found.

@@

@Suhler2011 also criticized the content and taxonomy of the five foundations. The authors criticize @Haidt2007's omissions of additional foundations, including *industry* and *modesty*, claiming these concepts are moralized in many societies worldwide. Likewise, the authors question whether or not the foundations are sufficiently distinct as to stand as their own foundation. For example, @Suhler2011 posit that *ingroup/loyalty* is merely a group-focused version of *harm/care*. Work by @Graham2011 might serve to rebut this criticism, as the researchers found their original five-factor structure seemed to best fit the data when validating the Moral Foundations Questionnaire. Finally, @Suhler2011 point out that particular concepts related to a foundation, including "anger" in *fairness/reciprocity* and "deception" in *ingroup/loyalty*, could be ascribed to any of the other foundations as well. In other words, it becomes difficult to recognize a particular concept as indicative of any one foundation when, in theory, they could be applied to all five [@Haidt2007; @Suhler2011].

These criticisms of Moral Foundations Theory are valid and should be taken into consideration when conducting research with instruments derived from MFT. The current authors argue these criticisms are especially valid when considered alongside questions regarding the Moral Foundations Dictionary they state herein. However, there exists compelling evidence regarding the validity of Moral Foundations Theory, albeit regarding its application solely through the Moral Foundations Questionnaire. This evidence is discussed within a brief explanation of the Moral Foundations Questionnaire and its relationship to the Moral Foundations Dictionary.

@@

The Moral Foundations Questionnaire has been validated by multiple researchers. Likewise, its five-factor structure has been demonstrated to fit data in multiple countries [@Graham2011; @Davies2014]. In their article introducing the 30-item MFQ, @Graham2011 conducted an exploratory factor analysis (EFA) and found that a five-factor model fit the data better than a one, two, or three-factor model. @Davies2014 conducted a confirmatory factor analysis (CFA) of the MFQ with a sample from New Zealand and likewise found the five-factor model provided the best fit. While @Davies2014 concede the US and New Zealand share many similarities as Western nations, which could raise questions regarding the validity of the MFQ in non-Western nations. However, there are striking differences between the two countries, including the lack of a two-party political system in New Zealand, that provide grounds for claiming the MFQ generalizes beyond the United States. Furthermore, @Graham2011 claim to find a "reasonable" degree of generalizability for the MFQ across participants from many different regions in the world. These two bodies of work also provide the best available evidence that the five moral foundations are sufficiently distinct from one another, though broader criticisms of MFT raised by @Suhler2011 should still be taken into account in studies involving the theory.

@@

While these concerns regarding the replication of social priming studies are valid and deserve further investigation, @Druckman2004 does not purport to demonstrate a widespread effect of social priming on the American electorate. In other words, this research makes no claim to empirically supported priming effects. Rather, @Druckman2004 chronicle the efforts on the part of the Nixon Administration to prop up the president's supposed strengths while downplaying his weaknesses. These tactics were deployed through the careful use of language in order to achieve the administration's political goals. As such, @Druckman2004's research on Nixon serves as an example of language's potential utility in the propagation of desirable political opinions. The researcher's investigation of news media's focus on specific issues during the Carter Administration likewise provide an example of language as a potential conduit for the transfer of politically biased information. The idea that even 1970s news media could contain political biases is of particular interest to the current study, which investigates similar phenomena in contemporary news media.

@@

For Experiment 1, the researchers approached the study with the intention to answer a method question. That is, this portion of the current research was conducted in order to solidify the best method by which to analyze political news text under the Moral Foundations Theory framework while also alleviating some of the aforementioned valence problem observed in the Moral Foundations Dictionary. The researchers hypothesized the news sources generally perceived as liberal leaning (*NPR* and *The New York Times*) would contain MFD words and valences indicating endorsements of the individualizing moral foundations (*harm/care* and *fairness/reciprocity*). Additionally, the researchers hypothesized the two sources generally perceived to be conservative leaning (*Fox News* and *Breitbart*) would feature MFD words and valences indicating equal endorsement of all five foundations. Owing to the lack of a need for human participants, the researchers did not petition Missouri State University's Institutional Review Board, as no such approval was needed to conduct this study.

@@

Political articles were collected from the websites of four notable U.S. news sources, a process known as web scraping. The sources were *The New York Times*, *National Public Radio (NPR)*, *Fox News*, and *Breitbart*. They were selected for their widespread recognition and the fact political partisans have strong preferences for some sources over others. The researchers determined the political lean of each source by referencing @Mitchell2014's article demonstrating the self-reported ideological consistency represented by the consumers of several news sources. In general, *The New York Times* and *NPR* are preferred by consumers reporting a liberal bias or lean. In contrast, *Fox News* and *Breitbart* are believed to have a conservative bias or lean. @Mitchell2014's article presented political ideology as a scale ranging from "consistently liberal" to "consistently conservative." In between these extremes lie more moderate positions, including "mostly liberal," "mixed," and "mostly conservative." Owing to the lower number of sources analyzed herein, the researchers elected to categorize the sources as either "liberal" and "conservative" in order to form a basis for comparison.

@@

The source article words were compiled into a dataset where they were matched up with their counterparts in the MFD along with their valence and a percentage of their occurrence. Therefore, for each article, the percentage of the number of *harm/care* words occurring in the articles were calculated, and this process was repeated for each of the foundations. Words' percent occurrence were multiplied by their *z*-scored valence. Valences were *z*-scored in order to eliminate any ambiguity regarding the direction of the valence. Positive values indicate positive valence, and negative values indicate negative valence. Words were categorized in accordance to their MFD affiliation, creating a weighted sum for each moral foundation.

@@

As seen in Table \@ref(tab:exp1-source-descriptives), the sources are similar in some aspects yet different in others. Valence appears to be slightly positive across all sources. The standard deviations seem to indicate little to no presence of a difference in valence. 
*The New York Times* published the greatest number of articles as well as total words. *Breitbart* featured the lowest number of articles, and *NPR* the lowest number of total words from all articles. Per individual article, however, *Breitbart* appears to feature the highest average number of words as well as unique words. Once again the standard deviations call into question any apparent differences between sources. Finally, *Fox News* appears to be the most readable source on average, while *The New York Times* is the least readable. This might be attributable to the greater number of tokens in the average *New York Times* article compared to *Fox News*. As before, the standard deviations for readability hamper one's ability to draw solid conclusions regarding differences in the sources' use of language.

@@

Owing to the exploratory nature of Experiment 1, the researchers were afforded the opportunity to consider changes to the method to be utilized in Experiment 2. Generally speaking, the researchers believe their methodology to be sound. Web scraping methods and text processing remain viable methods for collecting large amounts of text and subsequently rendering that text in a form suitable for data analysis. Experiment 1 also demonstrated a method by which to address inherent problems in the Moral Foundations Dictionary relating to valence. The solution provided in Experiment 1 appears to provide insights into the MFD words where none previously existed. Finally, calculating weighted percentages and sums for each moral foundation provides an easily interpreted summary of MFD word positivity and occurrence.

@@

The government shutdown ended 35 days later when Congress passed a continuing resolution to fund the government for three weeks while further negotiations regarding the budget would take place. President Trump signed the bill, ending the longest government shutdown in U.S. history [@Axelrod2019]. Many political commentators cite the mounting pressure on Trump and Congress stemming from significant costs both to unpaid American workers and the economy as a whole [@Kheel2019]. Perhaps owing to the financial pressures experienced by thousands of federal employees and their families as a result of a political quarrel, President Trump saw his approval rating fall as the shutdown progressed [@Peoples2019]. While the shutdown and its associated burdens were unpopular, the issues discussed concurrent to the shutdown offer the potential for more divergent opinions.

@@

In contrast to Experiment 1, the researchers approached Experiment 2 with the intention to confirm the method employed was valid for the analysis of the scraped text as well as for any inferences drawn from the analyses. For Experiment 2, the researchers hypothesized that news sources perceived as liberal will exhibit positive endorsements of the individualizing moral foundations (*harm/care* and *fairness/reciprocity*) in their articles reporting on both the Kavanaugh confirmation hearing as well as the 2018-2019 government shutdown. News sources perceived as conservative are hypothesized to positively endorse all five foundations equally in their coverage of the Kavanaugh hearing and the government shutdown. The researchers tested the hypothesis by analyzing the content scraped from news sources' web pages spanning the two weeks before (September 13, 2018) and two weeks after (October 11, 2018) Kavanaugh's confirmation hearing, owing to its prominence in the news. Likewise, the researchers analyzed content spanning two weeks before the start of the government shutdown (December 8, 2018) to two weeks following the end of the shutdown (February 8, 2019). The content will be analyzed for valence and moral alignment under Moral Foundations Theory. Once again, no human participants were needed for this study, so no Institutional Review Board approval was necessary.

@@

The same process was followed for scraping articles related to the partial U.S. Government shutdown of 2018-2019. The articles scraped were published between December 8, 2018 and February 8, 2019 inclusive. Once again, the researchers elected to scrape articles published two weeks before and after the event in question in order to capitalize on the shutdown's saturation in American news media.

@@

As in Experiment 1, the text was scanned with *ngram*. Again, blank articles, text from the Disqus system, and duplicate articles were removed [@Schmidt2017]. The text was processed and stemmed in order to convert to a usable form for further analysis [@Feinerer2017]. Words were subsequently matched with their valences from @Warriner2013. Depending on the results of @Jordan2019's research, alternative forms of the Moral Foundations Dictionary may be imported instead of the original dictionary.

Using the *tm* and *ngram* packages in *R*, the researchers processed the text in order to convert it to lowercase, fix spacing anomalies, and remove punctuation [@Feinerer2017]. Each individual word was reduced to its stem (i.e., *diseased* was stemmed to *diseas*). Once again, the same procedure was applied to the MFD words and the words in the @Warriner2013 dataset. Using the @Warriner2013 dictionary, the words in the MFD were assigned their respective valence. The researchers obtained the words' percent occurrence in the text. Once again, percents were multiplied by *z*-scored valence and categorized into their proper MFD category.

@@

Table \@ref(tab:exp2-source-descriptives-kav) displays the descriptive statistics for sources' writing on the Kavanaugh confirmation hearing. The sources were similar in most basic linguistic aspects, except for number of articles. For example, *The Huffington Post* appears to have published only 27 articles while *NPR* published 757 articles on this topic. Valence was found to be slightly positive across all sources. *Slate* produced the most total words with the most tokens on average. *Rush Limbaugh* featured the highest number of types on average. *The Blaze* was the most readable source on average by grade level while *Fox News* was the least readable. The standard deviations for these statistics, however, preclude conclusions regarding differences in the sources' use of language.

Table \@ref(tab:exp2-source-descriptives-gs) displays descriptive statistics for articles about the partial government shutdown. Like the Kavanaugh hearing, the sources were similar in average valence (slightly positive). Once again, there was variation in the number of articles published by each source on this topic. *The Hugginton Post* and *The Blaze* published fewer than 100 articles while *Slate* published 1,013 articles. *Slate* again featured the most total words and mean tokens. *Rush Limbaugh* again had the most types on average. For this topic, *Slate* was the most readable while *The Huffington Post* was the least readable. For each statistic, the standard deviations render any assertions regarding linguistic differences inconclusive on a descriptive level.

@@

To analyze if news sources adhered to differences in word use based on their target audience, the researchers utilized a multilevel model (MLM) to analyze if news sources leveraged different vocabularies based on target audience. Once again, the researchers used the *nlme* library in *R* [@Pinheiro2017]. Each foundation's weighted percentage was predicted by the source's political lean, using the individual source as a random intercept to control for the nested data structure. MLMs were constructed from datasets compiled for both topics of interest: the Kavanaugh hearing and the partial government shutdown of 2018-2019.

For the Kavanaugh topic, the multilevel model indicated the presence of a significant effect for *harm/care*, but the practical effect denoted by Cohen's *d* was found to be small. There were no other significant or practical effect of political lean for any of the other four moral foundations. The effect for *harm/care* was in the hypothesized direction with liberal sources tending to use more positively *harm/care* words than conservative sources. Descriptive and test statistics, *p*-values and effect sizes (Cohen's *d*) can be found in Table \@ref(tab:exp2-tablekav).

For news articles about the partial U.S. Federal Government Shutdown of 2018-2019, there were no significant or practical effects of political lean for the moral foundations. A small-to-medium effect size was observed for *authority/respect*. The effect was in the predicted direction as well, as conservative sources tended to offer more positive endorsements of the foundation than liberal sources. This is similar to Experiment 1 in which the largest effect size was observed for *authority/respect*. As noted before, the effect found in Experiment 1 was in the opposite direction as what was hypothesized. Thus, it is difficult to draw comparisons between the two studies despite the similar pattern for effect size. Owing to a lack of similar effects for either Experiment 1 or the Kavanaugh topic, there is doubt as to whether or not a practical or generalized effect exists for *authority/respect*.

@@

The results obtained in Experiment 2 did not confirm either of the hypotheses. The researchers found little compelling evidence of an effect of partisan lean on MFD endorsement. The only significant effect was found for *harm/care* in the Kavanaugh confirmation hearing dataset. While the effect was in the hypothesized direction (higher positive endorsement for liberal sources), the effect size was small. While there were no significant effects found for the government shutdown topic, a small-to-medium effect size was calculated for the *authority/respect* foundation. Once again, the effect was in the hypothesized direction with conservative sources providing higher positive endorsements of the foundation relative to liberal sources.

Both of these effects are consistent with both the research hypotheses as well as previous findings in Moral Foundations Theory scholarship. It should be noted these effects were found in isolation, and similar patterns were not found across the other studies. These findings undermine any inferences of political bias stemming from the content of news articles. The results obtained for Experiment 2 call into question the Moral Foundations Dictionary's efficacy as a tool for establishing differences between political news sources by partisan lean.

@@

Within the theoretical framework of Moral Foundations Theory [@Haidt2007], the researchers attempted to devise a method leveraging the Moral Foundations Dictionary [@Graham2009] in order to quantify political bias stemming from content published by several prominent American news sources. This method was carried out in several parts. First, the researchers collected the articles through web scraping techniques in *R* [@Wickham2016]. Second, the text was processed using *ngram* and *tm* in *R* to render it in a form conducive to computerized statistical analysis [@Feinerer2017]. Third, percent occurrence of words in the Moral Foundations Dictionary (MFD) was calculated on each source in order to establish how often words indicating endorsement of the five moral foundations appeared. Frequency of MFD word occurrence per article was weighted for valence by multiplying the scores by mean *z-scored* valence ratings from @Warriner2013. Finally, the researchers constructed multi-level models using *nlme* in *R* [@Pinheiro2017] for each foundation in which weighted percentages were predicted by each source's political lean. Individual source was held as a random intercept to control for correlated error. This process was undertaken once in Experiment 1 for general political news and twice in Experiment 2 using articles pertaining to two specific topics in American news: Brett Kavanaugh's confirmation to the Supreme Court and the most recent (as of this writing) partial U.S. Government shutdown from December 2018 to January 2019.

@@

For articles pertaining to the Kavanaugh hearing, there was a statistically significant effect for *harm/care*, but the effect size was modest. No other statistically or practically significant effects were found for the Kavanaugh topic. Likewise, there were no statistically or practically significant effects revealed for the government shutdown topic. However, there was a small-to-medium effect size for *authority/respect*. The fact these patterns do not appear in any of the other studies undertaken herein decreases the researchers' ability to draw any valuable conclusions regarding the MFD's ability to reveal political bias in American news media.

The results obtained in both experiments, while not confirmatory, provide several avenues for future investigation. Twice confronted with the failure to uncover differences in news media's political content, the researchers were unable to confirm the existence of political bias in American news under the auspices of Moral Foundations Theory. This does not, however, rule out the possibility that news media is biased. As @Mitchell2014 has demonstrated, self-reported political partisans are drawn to separate sources, but the present research was unable to offer any content-based explanations of this phenomenon. Even when attenuating the scraping criteria to collect text concerning (what the researchers believed to be) morally resonant topics, the method failed to detect differences in MFD word use by prominent news sources. The researchers' focus, therefore, turns toward the limitations of the method, especially the chosen theoretical framework, under which the present research was conducted.

Despite the fact the results regarding political bias were inconclusive, the researchers still retain confidence in the overall structure of the methodology established in the current study. Specifically, the procedure for scraping text from the web, processing, stemming, and weighting the scores with valence seems to represent a solid method for preparing a high quantity of text passages for data analysis. The researchers implemented valence as an indicator of the directionality of endorsement due to the inherent ambiguity of simply calculating MFD word percent occurrence. Additionally, multi-level models were necessary to control for the nested nature of the data where political lean encompassed several sources, which themselves encompassed numerous articles.

Thus, the researchers' attention turned to the instrument used to detect differences. In the present study, the researchers utilized the Moral Foundations Dictionary, a metric developed from Moral Foundations Theory [@Graham2009]. Based upon the results obtained, it might be necessary to investigate alternative instruments that might better elucidate the differences of interest. Likewise, other theoretical perspectives may be better equipped to explain political differences in discourse.

Tools such as these could be developed through the Linguistic Inquiry and Word Count program (LIWC) [@Pennebaker2007]. As stated before, the Moral Foundations Dictionary was established through an analysis of Christian sermons performed mostly with LIWC as well as some human-enforced corrections due to context [@Graham2009]. This approach was grounded in Moral Foundations Theory, a clear framework within which the authors developed the MFD. The authors established sound psychometric properties for the tool, implying its validity in several applications. For this reason, the researchers elected to utilize the MFD as their primary instrument.

Other instruments could theoretically be developed through a LIWC analysis, including those that would be useful for analyzing political discourse. Researchers could obtain percent occurrence scores for words appearing in certain LIWC categories, which could then be weighted with valence scores from @Warriner2013. The researchers will consider other applications of the LIWC in future studies aiming to analyze political content. However, without a firm theoretical foundation for such analyses, any such applications of LIWC would face serious questions as to their relevance, validity, and generalizability. Nevertheless, LIWC analyses may represent a path to further research into political differences in news media as well as theory formation.

There is, however, another theoretical framework that could prove useful when applied to the method established in the present research. Relational Frame Theory originates in applied behavior analysis, but there exists recent scholarship that demonstrates its potential usefulness in the political realm. Relational Frame Theory (RFT) holds that stimuli, including speech and discourse, elicit responses in terms of relationships to other stimuli, even when the stimuli do not otherwise appear to be related [@Blackledge2014]. For example, an individual's equivalent fear response to a venomous snake as well as a wooded area said to "contain snakes" represents "mutual entailment." That is, the fear response is an effect of both the snake and environment stimuli, which cause the fear response. "Combinatorial entailment" can be represented by a fear response elicited by a novel stimulus whose relationship to a well-known stimulus elicits a certain response [@Blackledge2014]. In the snake example, if another stimulus (Stimulus A), were said to be "worse than" the snake, the relationship between A and the snake would elicit a more severe fear response to Stimulus A. RFT utilizes a set of "families" describing the nature of these relationships. Such families include "coordinative," establishing similarities between stimuli, and "distinctive," implying fundamental differences between stimuli [@Belisle2018]. These families can be used to analyze networks of relational frames in discourse.
